Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1}

### 2024-08-19 doc/VerticalArrangements.tex
this version incorporates Gloria's edits to the intro and theory section. Includes changes to lit review and bibliography

\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}

@@

\begin{abstract}
\noindent We study the welfare impacts of mergers when some firms are already vertically integrated. Our model features logit Bertrand competition downstream and Nash Bargaining upstream.  We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs. In our model, we find that mergers with both horizontal and vertical characteristics typically harm consumers.  We apply the model to the Republic/Santek merger as a real-world example.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.  However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  As a result, the antitrust community has little consensus regarding how ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects, will affect welfare.

The horizontal versus vertical distinction has been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.  The DOJ did not.  }  Given these policy documents, it is not surprising that the agencies typically frame most mergers as either horizontal or vertical.  Similarly, most antitrust research maintains this separate categorization, leaving us with little systematic evidence for how complex mergers impact consumers.  However, as part of the current effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''  }  In this paper, we address this question using a vertical supply chain model to explore the welfare implications of complex mergers.

Our model is drawn from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  Relative to \citeN{ST2021}, the first category, vertical mergers, appears in the previous paper and here serves as a baseline for comparison.  The other three cases, which feature at least one merging firm that is already integrated, are new.

Our model-based simulations show certain patterns across mergers.  For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.  The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.  Likewise, we find that integrated mergers are typically net harmful to consumers as well.  The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.  Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.  We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results. We find that the existence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less harm.

We then employ this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee. The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.  Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.  In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.  Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.  We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.  Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts.

There are a few caveats to our analysis.  First, products in our merger simulations do not experience any changes in their qualities or costs, other than movements in their wholesale prices due to shifts in bargaining leverage.  Therefore, we rule out merger efficiencies as they are conceptualized in, for example, \citeN{werden96}, and we instead focus on EDM as the main channel through which mergers may benefit consumers.  Second, we model upstream bargaining as being over linear fees, rather than over two-part tariffs or other compensation schedules.  Although this assumption means bargaining is inefficient, linear input prices are widely observed in some industries, perhaps because of difficulties with implementing or monitoring more complicated contracts.\footnote{One example of an industry with linear wholesale pricing that has been featured in several vertical merger proceedings is cable programming distribution, as detailed in \citeN{Rogerson2014}.}  Third, our model assumes that upstream and downstream prices are set treating all other prices as fixed.  Although we make this assumption in part for simplicity, it also may be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.  In contrast, \citeN{BBMM2023} allow downstream firms to treat their own prices as adjustable during upstream negotiations.  The authors argue that this setup is useful for settings like grocery stores, where menu costs are low.  }  Fourth, we rely on the logit for our demand function, which limits the ways in which consumers substitute between products.  In \citeN{ST2012}, we found that assuming the merging firms operate in the same group within a nested logit model tends to increase the magnitude of consumer harm relative to the flat logit.

Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.  The closest merger simulation article is our previous work in \citeN{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger.  That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.  \citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.  \citeN{MS2013} and \citeN{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.  However, none of these papers study complex mergers.

Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \citeN{DKVB2010}.  These models have been applied to several sectors that are the frequent target of complex mergers, such as healthcare (e.g. \citeN{HL2017}) and video programming distribution (e.g. \citeN{CLWY2017}).  These structural empirical papers explore the welfare effects of various competitive interactions in certain industries.  Unlike these papers, our framework has been simplified for use with limited data, with the goal of being more accessible to practitioners.  These simplifications allow us to provide numerical results over a wide range of market conditions, which complements the insights that come from the structural empirical literature on specific case studies.

The paper proceeds as follows.  In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers.  Section \ref{sec:sims} explains how we construct our numerical simulations and provides our results, and our application to Republic/Santek appears in Section \ref{sec:application}.  Section \ref{sec:concl} concludes. Additional simulation results and information on the data used in the application appear in the Appendix.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory\label{sec:theory}}
We begin by describing our basic framework, taken from \citeN{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.  This framework allows us to study a variety of complex merger configurations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Downstream and Upstream Competition}
Assume there is a set of consumers who each choose to buy a single product sold by a retailer.  Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.  Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.  We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$ and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.  We divide the set $\mathbb{W}$ into overlapping subsets, each labeled $\mathbb{W}^r$, to indicate which wholesalers' products are carried by which retailers.  Similarly, we divide the set of retailers $\mathbb{R}$ into overlapping subsets, each labeled $\mathbb{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.

The share of consumers that choose product $w$ sold by retailer $r$ has the logit form,
\begin{equation}
s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{t \in \mathbb{R}} \sum_{x \in \mathbb{W}^t} \exp(\delta_{tx} - \alpha p_{tx})},
\label{eq: nb share}
\end{equation}
where $\delta_{rw}$ is a quality parameter and $\alpha$ captures sensitivity to price $p_{rw}$.  There is an outside good whose quality parameter and price have been normalized to zero.  The retailer's profit is given by $\pi^r = \sum_{w \in \mathbb{W}^r} [p_{rw} - p^W_{rw} - c^R_{rw}] s_{rw} M$, where $p^W_{rw}$ is the unit fee charged by wholesaler $w$ to retailer $r$, $c^R_{rw}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.  Downstream prices are set in Bertrand equilibrium, according to
\begin{equation}
\sum_{x \in \mathbb{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}
{\partial p_{rw}} + s_{rw} = 0,
\label{eq: nb downstream foc}
\end{equation}
which is the first order condition for product $w$ sold by retailer $r$.

Wholesale prices are set via Nash Bargaining between retailers and wholesalers.  We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.  }  Profits for wholesaler $w$ are given by $\pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M$, where $c^W_{rw}$ is the wholesale marginal cost when dealing with retailer $r$.  The first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
\begin{equation}
\begin{split}
&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right),
\end{split}
\label{eq: upstream foc}
\end{equation}
where $\lambda \in [0,1]$ captures the bargaining power of the retailer relative to the wholesaler.  The $\Delta s_{tx}(\mathbb{W}^r \setminus \{w\}) \equiv s_{tx}(\mathbb{W}^r \setminus \{w\}) - s_{tx}$ is the difference in the share of good $x$ sold by retailer $t$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.  Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.

Together the series of downstream and upstream first order conditions determine market equilibrium.  This model can be solved and calibrated as described in \citeN{ST2021}.  In our baseline configuration, we assume that wholesale and retail marginal costs are constant.  We allow for linear increasing marginal costs as a robustness check that appears in Appendix \ref{app:mc}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers}
Here we describe the manner in which complex mergers are modeled in this framework.  Throughout we assume that each product's quality, $\delta_{rw}$, and marginal costs, $c^R_{rw}$ and $c^W_{rw}$, remain unchanged after the merger.  We begin by describing vertical mergers, as all cases we examine have some vertical aspects.  Suppose that retailer $r$ and wholesaler $w$ were to merge.  Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $rw$ is given by
\begin{equation}
\begin{split}
&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{EDM effect}} \\
&+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} = 0.
\end{split}
\label{eq: vmerger own ret downstream foc}
\end{equation}
The pricing problem balances two effects.  On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $r$ being able to access product $w$ at marginal cost.  This force would tend to lower the resulting price.  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to wholesaler $w$.  This force would tend to raise the resulting consumer price.

Turning to input prices, when wholesaler $w$ bargains with unaffiliated retailer $s$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
\end{split}
\label{eq: vmerger wh upstream foc}
\end{equation}
which reflects the change in the disagreement payoff coming from the merger with retailer $r$.  Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $s$, these losses are softened due to a potential for diversion to retailer $r$, which we label the ``RRC effect.''  Furthermore, the margin on product $w$ sold by retailer $r$ is potentially higher due to EDM, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.  These impacts tend to raise the resulting input price.

>>>>

\title{Beyond Horizontal and Vertical:\\ Simulating the Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}

@@

\begin{abstract}
\noindent We study the simulation of mergers that have both horizontal and vertical aspects. Our model features logit Bertrand competition downstream and Nash Bargaining upstream.  We apply the model to the Republic/Santek merger as a real-world example.  %We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs. We apply the model to the Republic/Santek merger as a real-world example.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely horizontal or purely vertical.  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.  However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  As a result, the antitrust community has done little research into ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects.

The horizontal versus vertical distinction had been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The most recent Horizontal Merger Guidelines is from 2010, while the most recent Vertical Merger Guidelines is from 2020.  They have both been recently superseded by a joint document.  }  Given these policy documents, it is not surprising that the agencies have historically framed most mergers as either horizontal or vertical.  Similarly, most antitrust research maintains this separate categorization, leaving us with little guidance for how our merger simulations and related tools apply to complex mergers. However, in recognition of these issues, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical aspects of competition.  As part of the revision effort, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''  }  In this paper, we address this question by exploring the implications that a typical merger simulation model has for the welfare effects of complex mergers.

%We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  Relative to \citeN{ST2021}, the first category, vertical mergers, appears in the previous paper and here serves as a baseline for comparison.  The other three cases, which feature at least one merging firm that is already integrated, are new.

%Our model-based simulations show certain patterns across mergers.  For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.  The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.  Likewise, we find that integrated mergers are typically net harmful to consumers as well.  The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.  Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.  We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results. We find that the existence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less harm.

Our model is drawn from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee. The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.  Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.  In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.  Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.  We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.  Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts.

There are a few caveats to our analysis.  First, products in our merger simulations do not experience any changes in their qualities or costs, other than movements in their wholesale prices due to shifts in bargaining leverage.  Therefore, we rule out merger efficiencies as they are conceptualized in, for example, \citeN{werden96}, and we instead focus on EDM as the main channel through which mergers may benefit consumers.  Second, we model upstream bargaining as being over linear fees, rather than over two-part tariffs or other compensation schedules.  Although this assumption means bargaining is inefficient, linear input prices are widely observed in some industries, perhaps because of difficulties with implementing or monitoring more complicated contracts.\footnote{One example of an industry with linear wholesale pricing that has been featured in several vertical merger proceedings is cable programming distribution, as detailed in \citeN{Rogerson2014}.}  Third, our model assumes that upstream and downstream prices are set treating all other prices as fixed.  Although we make this assumption in part for simplicity, it also may be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.  In contrast, \citeN{BBMM2023} allow downstream firms to treat their own prices as adjustable during upstream negotiations.  The authors argue that this setup is useful for settings like grocery stores, where menu costs are low.  }  Fourth, we rely on the logit for our demand function, which limits the ways in which consumers substitute between products.  We also provide additional results using the nested logit.

Our paper is related to the large literature on competition and vertical integration that grew out of a number of theory articles in the 1980s and 1990s.  \citeN{Riordan2008} has an excellent summary. This research explores how incentives to disadvantage rivals could arise alongside the efficiencies (such as EDM) associated with vertical integration.  Some of these incentives result in RRC effects (e.g. \citeN{SalopScheffman1983} or \citeN{Riordan1998}) or in the foreclosure of competitors from input markets (e.g. \citeN{Salinger1988} or \citeN{OSS1990}).  Unlike in our paper, most of these models are not framed explicitly as bargaining problems, instead assuming firms making take-it-or-leave-it offers.  However, many of the incentives towards RRC, which usually involve trading off profits from input sales against those from downstream sales, are similar to those that appear in our model. These theory papers tend to focus on pure vertical mergers rather than complex mergers, in order to maintain mathematical tractability.

Meanwhile, the structural empirical industrial organization literature has also studied the competitive effects of vertical integration as part of a greater exploration of the impacts of bargaining power in vertical supply chains. Many of these papers are based on the model developed in \citeN{HW1988} modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.  These models have been applied to several sectors that are the frequent target of vertical integration, such as healthcare (e.g. \citeN{HL2017}) and video programming distribution (e.g. \citeN{CLWY2017}).  These structural empirical papers explore the welfare effects of various competitive interactions, including complex mergers, in certain industries.\footnote{There is a companion empirical literature that relies more on regression analysis that identified changes in market outcomes due to vertical integration (e.g. \citeN{Chipty2001} on video programming distribution).  }  Our framework is similar to the models appearing in these papers, but it has been simplified for use with limited data, with the goal of being more accessible to practitioners and policymakers.

Our work is also related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.  The closest merger simulation article is our previous research in \citeN{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger, ruling out complex mergers.  That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.  \citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.  \citeN{MS2013}, \citeN{Rogerson2020}, and \citeN{DSS2024} develop UPP-style tools for assessing the impact of vertical mergers.  However, these papers do not focus on complex mergers.

The paper proceeds as follows.  In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers.  Our application to Republic/Santek appears in Section \ref{sec:application}.  Section \ref{sec:concl} concludes. Additional simulation results and information on the data used in the application appear in the Appendix.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory\label{sec:theory}}
We begin by describing our basic framework, based on \citeN{ST2021}, which uses a downstream Bertrand model embedded in an upstream Nash Bargaining setup.  We extend this framework allows to study a variety of complex merger configurations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Downstream and Upstream Competition}
Assume there is a set of consumers who each choose to buy a single product sold by a retailer.  Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.  Prior to any mergers taking place, each wholesaler offers only one product, although each retailer may purchase from multiple wholesalers.  Therefore, unique products (meaning a wholesale product offered at a given retail outlet) are denoted by pairs of the retailer and wholesaler indices, $rw$.  We denote the set of all retailers by $\mathcal{R}=\{1, \dots, \left\vert{\mathcal{R}}\right\vert\}$ and the set of all wholesalers by $\mathcal{W}=\{1, \dots, \left\vert{\mathcal{W}}\right\vert\}$.  We divide the set $\mathcal{W}$ into potentially overlapping subsets, each labeled $\mathcal{W}^r$, to indicate which wholesalers' products are carried by which retailers.  Similarly, we divide the set of retailers $\mathcal{R}$ into potentially overlapping subsets, each labeled $\mathcal{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.  Note that throughout we take as given the set of retailer-wholesaler combinations being offered in the market.  We do not model the process of firm entry and exit or the way in which retailers choose which wholesale products to offer.

Let the share of consumers that choose product $w$ sold by retailer $r$ be denoted by $s_{rw}$.  This share is a function of the vector of retail prices $P$ offered across all available products.\footnote{We omit the retail price argument for $s_{rw}(P)$ in what follows for brevity's sake.}  In our simulations we model the share using both logit and nested logit functions.  The random coefficients logit is another commonly used option.  The profit for retailer $r$ is given by $\pi^r = \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx} M$, where $p^W_{rx}$ is the unit fee charged by wholesaler $x$ to retailer $r$, $c^R_{rx}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.  Downstream prices are set in Bertrand equilibrium according to
\begin{equation}
\sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}
{\partial p_{rw}} + s_{rw} = 0,
\label{eq: nb downstream foc}
\end{equation}
which is the first order condition for product $w$ sold by retailer $r$.

Wholesale prices are set via bilaterial Nash Bargaining between retailers and wholesalers.  We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.  }  Profits for wholesaler $w$ are given by $\pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M$, where $c^W_{xw}$ is the wholesale marginal cost when dealing with retailer $x$.  The first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
\begin{equation}
\begin{split}
&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right),
\end{split}
\label{eq: upstream foc}
\end{equation}
where $\lambda \in [0,1]$ captures the bargaining power of the retailer relative to the wholesaler.  The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) \equiv s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.  Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.

Together the series of downstream and upstream first order conditions determine market equilibrium.  After assuming a functional form for shares, the model can be solved and calibrated along the lines described in \citeN{ST2021}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers}
Here we describe the manner in which complex mergers are modeled in this framework.  Throughout we assume that all marginal costs, $c^R_{rw}$ and $c^W_{rw}$, and any non-price product characteristics remain unchanged after the merger, which rules out merger efficiencies.  We begin by describing vertical mergers, as all cases we examine have some vertical aspects.  Suppose that retailer $t$ and wholesaler $v$ were to merge.  Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $tv$ is given by
\begin{equation}
\begin{split}
&\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} + s_{tv} + \overbrace{[p_{tv} - c^W_{tv} - c^R_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}^{\text{EDM effect}} \\
&+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_{\text{upstream UPP effect}} = 0.
\end{split}
\label{eq: vmerger own ret downstream foc}
\end{equation}
The pricing problem balances two effects.  On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $t$ being able to access product $v$ at marginal cost.  This force would tend to lower the resulting price.\footnote{However, this lower marginal cost raises the potential for a larger margin, which could also cause a multi-product retailer to raise prices on its other goods $x \in \mathcal{W}^t \setminus \{v\}$, in order to divert sales to product $tv$.  This is the Edgeworth-Salinger effect as in \citeN{Salinger1991}.}  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $t$ in order to divert sales to wholesaler $v$.  This force would tend to raise the resulting consumer price.
